---
title: 'Houston Blackouts'
author: "Charlie Curtin"
date: "2023-12-15"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE)

options(scipen = 999) # disable scientific notation
```

## Overview

"In February 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10--11, 13--17, and 15--20."[^1] For more background, check out these [engineering](https://www.youtube.com/watch?v=08mwXICY4JM&ab_channel=PracticalEngineering) and [political](https://www.youtube.com/watch?v=Zcrsgdl_hP0&ab_channel=Vox) perspectives.

[^1]: Wikipedia. 2021. "2021 Texas power crisis." Last modified October 2, 2021. <https://en.wikipedia.org/wiki/2021_Texas_power_crisis>.

In this analysis, we will:\
- estimate the number of homes in Houston that lost power as a result of the first two storms\
- investigate if socioeconomic factors are predictors of communities recovery from a power outage

We will be using remotely-sensed night lights data, acquired from the [Visible Infrared Imaging Radiometer Suite (VIIRS)](https://en.wikipedia.org/wiki/Visible_Infrared_Imaging_Radiometer_Suite) onboard the Suomi satellite. In particular, we will use the VNP46A1 to detect differences in night lights before and after the storm to identify areas that lost electric power.

To determine the number of homes that lost power, we'll link our night lights data with Open Street Maps roads and Texas census tracts.

### Data

#### Night lights

We'll use NASA's Worldview to explore the data around the day of the storm. There are several days with too much cloud cover to be useful, but 2021-02-07 and 2021-02-16 provide two clear, contrasting images to visualize the extent of the power outage in Texas.

VIIRS data is distributed through NASA's [Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC)](https://ladsweb.modaps.eosdis.nasa.gov/). Many NASA Earth data products are distributed in 10x10 degree tiles in sinusoidal equal-area projection. Tiles are identified by their horizontal and vertical position in the grid. Houston lies on the border of tiles h08v05 and h08v06. We therefore need to download two tiles per date.

#### Roads

Typically highways account for a large portion of the night lights observable from space (see Google's [Earth at Night](https://earth.google.com/web/@27.44405464,-84.7693044,206.63660162a,8916361.52264659d,35y,0h,0t,0r/data=CiQSIhIgMGY3ZTJkYzdlOGExMTFlNjk5MGQ2ZjgxOGQ2OWE2ZTc)). To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas near highways.

[OpenStreetMap (OSM)](https://planet.openstreetmap.org/) is a collaborative project which creates publicly available geographic data of the world. Ingesting this data into a database where it can be subsetted and processed is a large undertaking. Fortunately, third party companies redistribute OSM data. We used [Geofabrik's download sites](https://download.geofabrik.de/) to retrieve a shapefile of all highways in Texas and prepared a Geopackage (`.gpkg` file) containing just the subset of roads that intersect the Houston metropolitan area. 

#### Houses

We can also obtain building data from OpenStreetMap. We again downloaded from Geofabrick and prepared a GeoPackage containing only houses in the Houston metropolitan area.\

#### Socioeconomic

We cannot readily get socioeconomic information for every home, so instead we obtained data from the [U.S. Census Bureau's American Community Survey](https://www.census.gov/programs-surveys/acs) for census tracts in 2019. The *folder* `ACS_2019_5YR_TRACT_48.gdb` is an ArcGIS ["file geodatabase"](https://desktop.arcgis.com/en/arcmap/latest/manage-data/administer-file-gdbs/file-geodatabases.htm), a multi-file proprietary format that's roughly analogous to a GeoPackage file.\

## Assignment

Below is an outline of the analysis steps we take.

```{r}
# import the necessary packages
library(stars)
library(sf)
library(tidyverse)
library(ggplot2)
library(ggspatial)
library(terra)
library(raster)
```

#### Finding locations of blackouts

For improved computational efficiency and easier interoperability with `sf`, we'll use the `stars` package for raster handling.\

##### Combining the data

First, we'll read in the night lights tiles and combine them into a single stars object for each date (2021-02-07 and 2021-02-16) using `st_mosaic`.

```{r include=TRUE}
# read in the night lights tiles as stars objects
h08v05_02_07 <- read_stars('data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif')

h08v06_02_07 <- read_stars('data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif')

h08v05_02_16 <- read_stars('data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif')

h08v06_02_16 <- read_stars('data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif')

# create a mosaic of night lights tiles for each date
night_lights_02_07 <- st_mosaic(h08v05_02_07, h08v06_02_07)

night_lights_02_16 <- st_mosaic(h08v05_02_16, h08v06_02_16)
```

##### Creating and vectorizing a blackout mask

To find the change in night lights intensity (presumably) caused by the storm, we'll reclassify the difference raster, assuming that any location that experienced a drop of more than 200 nW cm^-2^sr^-1^ experienced a blackout. We'll assign `NA` to all locations that experienced a drop of *less* than 200 nW cm^-2^sr^-1^. Then, we'll vectorize the blackout mask using `st_as_sf()`. We can use `st_make_valid()` to fix any invalid geometries.

```{r include=TRUE}
# find the difference raster
# we subtract the values on the 16th (when the storm was) from those on the 7th to find drops of light intensity 
light_diff_rast <- night_lights_02_07 - night_lights_02_16

# reclassify the difference raster to show blackout areas
light_diff_rast[light_diff_rast <= 200] = NA 

# vectorize the blackout mask and make the geometries valid
light_diff_vector <- light_diff_rast %>% 
  st_as_sf() %>% 
  st_make_valid() 
```

##### Cropping the vectorized map to the Houston Metropolitan area.

We can crop our vectorized blackout mask to the Houston metro area by passing these coordinates as a bounding box:\
- (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)\

Then we'll reproject the cropped blackout mask to EPSG:3083 to match the coordinate reference system of the census tract data we are going to use. Since we know that highways account for a great deal of light, we can exclude them from our blackout mask. We'll use a SQL query to select highways from the OpenStreetMaps geopackage. Then, we'll use `st_difference()` to exclude highways from the blackout mask.

```{r include=TRUE}
# create a bounding box for the Houston metro area
# create a matrix of coordinates first, repeating the first pair of coordinates to close the box
coords <- matrix(c(-96.5, 29, -96.5, 30.5, -94.5, 30.5, -94.5, 29, -96.5, 29), 
                 ncol = 2, # create two columns for the X and Y values
                 byrow = TRUE) # fill by row so the coordinate pairs are matched in rows

# create a polygon from the coordinates 
houston_metro <- st_polygon(list(coords)) %>% # pass the coordinates matrix as a list
  st_sfc(crs = st_crs(light_diff_vector)) # match the crs to our light difference vector

# crop our blackout mask to the houston metro area using spatial subsetting
houston_blackout <- light_diff_vector[houston_metro, ] %>% 
  st_transform(crs = 3083) # reproject to Texas Centric Albers Equal Area

# define our query to select only highways
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass = 'motorway'"

# read in only Texas highways from the roads geopackage
highways <- st_read("data/gis_osm_roads_free_1.gpkg", 
                    query = query,
                    quiet = TRUE) %>% 
  st_transform(crs = 3083)

# create a 200m buffer around highways, building in a check to ensure that our buffer distance is in meters
if (st_crs(highways)$units_gdal == "metre") { # ensure that the CRS units are meters
  highways_buffer <- highways %>% 
  st_buffer(dist = 200) %>% 
  st_union()
} else {
  print("Check the units in the CRS")
}

# exclude highways from the blackout mask by using st_difference
houston_blackout <- st_difference(houston_blackout, highways_buffer)
```

#### Finding homes impacted by blackouts

To find homes impacted by blackouts, we'll query residential buildings from the OpenStreetMaps buildings geopackage. We'll use spatial subsetting to filter to homes within blackout areas.

```{r include=TRUE}
# define our query to select only residential buildings
query2 <- "SELECT*
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

# select residences from our buildings geopackage using our query and reproject to Texas Centric Albers Equal Area
houston_houses <- st_read('data/gis_osm_buildings_a_free_1.gpkg',
                          quiet = TRUE,
                          query = query2) %>%
  st_transform(crs = 3083)

# filter to homes within blackout areas using spatial subsetting
blackout_houses <- houston_houses[houston_blackout, ]

# print the number of homes in blackout areas, building in a check to ensure that it is a subset
if (nrow(blackout_houses) < nrow(houston_houses)) {
  print(paste("Number of houston homes in blackout areas:", length(unique(blackout_houses$osm_id))))
} else {
  print("Check your subsetting method")
}
```

#### Investigating socioeconomic factors

In this section, we'll use the American Community Survey data to determine the socieconomic landscape of census tracts that experienced blackouts. We'll join the tabular income data to the census tract geometries by census tract ID. Then, we'll use `st_intersects` with impacted homes and census tracts to find out which census tracts experienced blackouts.

```{r include=TRUE}
# read in Texas census tracts data and reproject to Texas Centric Albers Equal Area
texas_tracts <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", 
                  layer = "ACS_2019_5YR_TRACT_48_TEXAS", 
                  quiet = TRUE) %>% 
  st_transform(crs = 3083)

# read in the Texas census tracts tabular data and select the median income and GEOID columns
tracts_income <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", 
                  layer = "X19_INCOME", 
                  quiet = TRUE) %>% 
  dplyr::select(B19013e1, GEOID)

# merge the census tracts and income data by common columns
texas_tracts <- merge(texas_tracts, tracts_income, by.x = "GEOID_Data", by.y = "GEOID")

# use st_intersects to find which census tracts experienced blackouts
texas_tracts$blackout <- lengths(st_intersects(texas_tracts, blackout_houses)) > 0

# crop census tracts to the houston metro area
houston_tracts <- st_crop(texas_tracts, st_bbox(houston_blackout))

# print the number of impacted census tracts
print(paste("The number of Houston census tracts impacted by blackouts:",
            nrow(filter(houston_tracts, blackout == TRUE))))
```

##### Comparing incomes of impacted tracts to unimpacted tracts

Now let's compare the median income of impacted tracts to unimpacted tracts. We'll create a map of median income by census tracts, overlaying red boundaries on census tracts that had blackouts. Then we'll create a boxplot of the distribution of median income in impacted and unimpacted census tracts.

```{r}
# filter and subset for census tracts that experienced a blackout
blackout_tracts <- houston_tracts %>% 
  filter(blackout == TRUE)

# map census tracts by median income, highlighting census tracts that experienced a blackout
blackout_map <- ggplot() +
  geom_sf(data = houston_tracts, # fill census tracts by median income and add a white border
          aes(fill = B19013e1),
          lwd = .1,
          col = "white") +
  scale_fill_viridis_b(name = "median income") + # add a name to the legend items
  geom_sf(data = blackout_tracts, # color borders of impacted tracts red
          aes(col = blackout),
          fill = "transparent",
          lwd = .2) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "transparent"),
                     breaks = "TRUE", # only include the symbol for impacted tracts in the legend
                     labels = c("experienced blackout"), # change the legend item label
                     name = "") +
  theme_void() + # remove latitude and longitude ticks
  labs(title = "Median Income and Blackouts by Census Tract in Houston") + # update the map title
  annotation_scale(width_hint = .4,
                   line_col = "white", 
                   text_col = "white",
                   pad_y = unit(.25, "in"),
                   pad_x = unit(.2, "in")) + # add a scale bar and change the color to white
  annotation_north_arrow(style = north_arrow_fancy_orienteering(line_col = "white", 
                                                                text_col = "white"),
                         pad_y = unit(.4, "in")) # add a north arrow and change the color to white

# save output map to images folder
ggsave(file.path("images", "blackout_map.jpg"), 
       blackout_map,
       width = 8, 
       height = 6, 
       units = "in", 
       dpi = 300)

# view the map
blackout_map
```

```{r}
# find the median median income by blackout status
medians <- houston_tracts %>% 
  group_by(blackout) %>% 
  summarize(median = median(B19013e1, na.rm = TRUE))

# visualize the distribution of income in impacted versus unimpacted census tracts using a boxplot
median_income <- ggplot(houston_tracts, aes(x = B19013e1, y = blackout, fill = blackout)) +
  geom_boxplot() +
  geom_text(data = medians, # add median labels for each group
            aes(x = median, y = blackout, label = median), 
            size = 3, 
            vjust = -.5, # move the label off of the line
            angle = -90) + # rotate the label
  theme_bw() +
  theme(legend.position = "none") + # remove the legend
  labs(title = "Distribution of Median Income in Census Tracts by Blackout Status",
       x = "median income",
       y = "blackout status") + # update the title and x and y axes labels
  scale_y_discrete(labels = c("no blackout", "blackout")) # update the blackout group labels

# save the output plot
ggsave(file.path("images", "median_income.png"), 
       median_income,
       width = 8, 
       height = 6, 
       units = "in", 
       dpi = 300)

# view the output plot
median_income
```

##### Discussion

The box plot displaying the distribution of median income in impacted and unimpacted census tracts shows that impacted census tracts actually had a slightly higher median median income than unimpacted census tracts. The difference is small, with impacted census tracts showing a median median income of \$ `r medians$median[2]` versus unimpacted census tracts showing a median median income of \$ `r medians$median[1]` . While we might expect to see a lower median median income in impacted census tracts, that isn't the case. A big limitation of our study could be the availability and accuracy of our light intensity rasters. The rasters appeared to have noise that could have affected the creation of our blackout mask, especially since we manually assigned the blackout threshold to be a difference of 200 or more nW cm^-2^sr^-1^ . Without accounting for noise, we may have misrepresented blackout areas.
