---
title: 'Houston Blackouts'
author: "Charlie Curtin"
date: "2023-12-15"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE)

options(scipen = 999) # disable scientific notation
```

## Overview

"In February 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10--11, 13--17, and 15--20."[^1] For more background, check out these [engineering](https://www.youtube.com/watch?v=08mwXICY4JM&ab_channel=PracticalEngineering) and [political](https://www.youtube.com/watch?v=Zcrsgdl_hP0&ab_channel=Vox) perspectives.

[^1]: Wikipedia. 2021. "2021 Texas power crisis." Last modified October 2, 2021. <https://en.wikipedia.org/wiki/2021_Texas_power_crisis>.

For this assignment, you are tasked with:\
- estimating the number of homes in Houston that lost power as a result of the first two storms\
- investigating if socioeconomic factors are predictors of communities recovery from a power outage

Your analysis will be based on remotely-sensed night lights data, acquired from the [Visible Infrared Imaging Radiometer Suite (VIIRS)](https://en.wikipedia.org/wiki/Visible_Infrared_Imaging_Radiometer_Suite) onboard the Suomi satellite. In particular, you will use the VNP46A1 to detect differences in night lights before and after the storm to identify areas that lost electric power.

To determine the number of homes that lost power, you link (spatially join) these areas with [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84) data on buildings and roads.

To investigate potential socioeconomic factors that influenced recovery, you will link your analysis with data from the US Census Bureau.

##### Learning objectives:

-   load vector/raster data\
-   simple raster operations\
-   simple vector operations\
-   spatial joins

### Data

#### Night lights

Use NASA's Worldview to explore the data around the day of the storm. There are several days with too much cloud cover to be useful, but 2021-02-07 and 2021-02-16 provide two clear, contrasting images to visualize the extent of the power outage in Texas.

VIIRS data is distributed through NASA's [Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC)](https://ladsweb.modaps.eosdis.nasa.gov/). Many NASA Earth data products are distributed in 10x10 degree tiles in sinusoidal equal-area projection. Tiles are identified by their horizontal and vertical position in the grid. Houston lies on the border of tiles h08v05 and h08v06. We therefore need to download two tiles per date.

As you're learning in EDS 220, accessing, downloading, and preparing remote sensing data is a skill in it's own right! To prevent this assignment from being a large data wrangling challenge, we have downloaded and prepped the following files for you to work with, stored in the `VNP46A1` folder.\

-   `VNP46A1.A2021038.h08v05.001.2021039064328.h5.tif`: tile h08v05, collected on 2021-02-07\
-   `VNP46A1.A2021038.h08v06.001.2021039064329.h5.tif`: tile h08v06, collected on 2021-02-07\
-   `VNP46A1.A2021047.h08v05.001.2021048091106.h5.tif`: tile h08v05, collected on 2021-02-16\
-   `VNP46A1.A2021047.h08v06.001.2021048091105.h5.tif`: tile h08v06, collected on 2021-02-16

#### Roads

Typically highways account for a large portion of the night lights observable from space (see Google's [Earth at Night](https://earth.google.com/web/@27.44405464,-84.7693044,206.63660162a,8916361.52264659d,35y,0h,0t,0r/data=CiQSIhIgMGY3ZTJkYzdlOGExMTFlNjk5MGQ2ZjgxOGQ2OWE2ZTc)). To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas near highways.

[OpenStreetMap (OSM)](https://planet.openstreetmap.org/) is a collaborative project which creates publicly available geographic data of the world. Ingesting this data into a database where it can be subsetted and processed is a large undertaking. Fortunately, third party companies redistribute OSM data. We used [Geofabrik's download sites](https://download.geofabrik.de/) to retrieve a shapefile of all highways in Texas and prepared a Geopackage (`.gpkg` file) containing just the subset of roads that intersect the Houston metropolitan area. 

-   `gis_osm_roads_free_1.gpkg`

#### Houses

We can also obtain building data from OpenStreetMap. We again downloaded from Geofabrick and prepared a GeoPackage containing only houses in the Houston metropolitan area.\

-   `gis_osm_buildings_a_free_1.gpkg`

#### Socioeconomic

We cannot readily get socioeconomic information for every home, so instead we obtained data from the [U.S. Census Bureau's American Community Survey](https://www.census.gov/programs-surveys/acs) for census tracts in 2019. The *folder* `ACS_2019_5YR_TRACT_48.gdb` is an ArcGIS ["file geodatabase"](https://desktop.arcgis.com/en/arcmap/latest/manage-data/administer-file-gdbs/file-geodatabases.htm), a multi-file proprietary format that's roughly analogous to a GeoPackage file.\

You can use `st_layers()` to explore the contents of the geodatabase. Each layer contains a subset of the fields documents in the [ACS metadata](https://www2.census.gov/geo/docs/maps-data/data/tiger/prejoined/ACSMetadata2011.txt).\

The geodatabase contains a layer holding the geometry information, separate from the layers holding the ACS attributes. You have to combine the geometry with the attributes to get a feature layer that `sf` can use.

## Assignment

Below is an outline of the steps you should consider taking to achieve the assignment tasks.

```{r}
# import the necessary packages
library(stars)
library(sf)
library(tidyverse)
library(ggplot2)
library(ggspatial)
library(terra)
library(raster)
```

#### Find locations of blackouts

For improved computational efficiency and easier interoperability with `sf`, I recommend using the `stars` package for raster handling.\

##### combine the data (5 points)

-   read in night lights tiles\
-   combine tiles into a single `stars` object for each date (2021-02-07 and 2021-02-16)\
    -   hint: use `st_mosaic`\

```{r include=TRUE}
# read in the night lights tiles as stars objects
h08v05_02_07 <- read_stars('data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif')

h08v06_02_07 <- read_stars('data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif')

h08v05_02_16 <- read_stars('data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif')

h08v06_02_16 <- read_stars('data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif')

# create a mosaic of night lights tiles for each date
night_lights_02_07 <- st_mosaic(h08v05_02_07, h08v06_02_07)

night_lights_02_16 <- st_mosaic(h08v05_02_16, h08v06_02_16)
```

##### create a blackout mask (10 points)

-   find the change in night lights intensity (presumably) caused by the storm\
-   reclassify the difference raster, assuming that any location that experienced a drop of more than 200 nW cm^-2^sr^-1^ experienced a blackout\
-   assign `NA` to all locations that experienced a drop of *less* than 200 nW cm^-2^sr^-1^\

```{r include=TRUE}
# find the difference raster
# we subtract the values on the 16th (when the storm was) from those on the 7th to find drops of light intensity 
light_diff_rast <- night_lights_02_07 - night_lights_02_16

# reclassify the difference raster to show blackout areas
light_diff_rast[light_diff_rast <= 200] = NA 
```

##### vectorize the mask (5 points)

-   use `st_as_sf()` to vectorize the blackout mask\
-   fix any invalid geometries using `st_make_valid`

```{r include=TRUE}
# vectorize the blackout mask and make the geometries valid
light_diff_vector <- light_diff_rast %>% 
  st_as_sf() %>% 
  st_make_valid() 

# check that only our blackout areas are included
if (min(light_diff_vector$VNP46A1.A2021038.h08v05.001.2021039064328.tif) >= 201) { # only values of light difference 201 or greater should be included
  print("Only blackout areas are included in our vector")
} else {
  print("Check your difference raster and your vectorization again!")
}
```

##### crop the vectorized map to our region of interest (10 points)

-   define the Houston metropolitan area with the following coordinates\
    -   (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)\
-   turn these coordinates into a polygon using `st_polygon`\
-   convert the polygon into a simple feature collection using `st_sfc()` and assign a CRS\
    -   hint: because we are using this polygon to crop the night lights data it needs the same CRS\
-   crop (spatially subset) the blackout mask to our region of interest 
-   re-project the cropped blackout dataset to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)\

```{r include=TRUE}
# create a bounding box for the Houston metro area
# create a matrix of coordinates first, repeating the first pair of coordinates to close the box
coords <- matrix(c(-96.5, 29, -96.5, 30.5, -94.5, 30.5, -94.5, 29, -96.5, 29), 
                 ncol = 2, # create two columns for the X and Y values
                 byrow = TRUE) # fill by row so the coordinate pairs are matched in rows

# create a polygon from the coordinates 
houston_metro <- st_polygon(list(coords)) %>% # pass the coordinates matrix as a list
  st_sfc(crs = st_crs(light_diff_vector)) # match the crs to our light difference vector

# crop our blackout mask to the houston metro area using spatial subsetting
houston_blackout <- light_diff_vector[houston_metro, ] %>% 
  st_transform(crs = 3083) # reproject to Texas Centric Albers Equal Area

# check that our cropping worked
if (nrow(houston_blackout) < nrow(light_diff_vector)) {
  print(paste("The blackout mask was cropped from", nrow(light_diff_vector), "to", nrow(houston_blackout), "polygons"))
} else {
  print("The blackout mask was not cropped correctly")
}
```

##### exclude highways from blackout mask (10 points)

The roads geopackage includes data on roads other than highways. However, we can avoid reading in data we don't need by taking advantage of `st_read`'s ability to subset using a SQL query.\

-   define SQL query\
-   load just highway data from geopackage using `st_read`\
-   reproject data to EPSG:3083\
-   identify areas within 200m of all highways using `st_buffer`\
    -   hint: `st_buffer` produces undissolved buffers, use `st_union` to dissolve them\
-   find areas that experienced blackouts that are further than 200m from a highway

`query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"`\
`highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query)`

```{r include=TRUE}
# define our query to select only highways
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass = 'motorway'"

# read in only Texas highways from the roads geopackage
highways <- st_read("data/gis_osm_roads_free_1.gpkg", 
                    query = query,
                    quiet = TRUE) %>% 
  st_transform(crs = 3083)

# create a 200m buffer around highways, building in a check to ensure that our buffer distance is in meters
if (st_crs(highways)$units_gdal == "metre") { # ensure that the CRS units are meters
  highways_buffer <- highways %>% 
  st_buffer(dist = 200) %>% 
  st_union()
} else {
  print("Check the units in the CRS")
}

# exclude highways from the blackout mask by using st_difference
houston_blackout <- st_difference(houston_blackout, highways_buffer)
```

#### Find homes impacted by blackouts

##### load buildings data (10 points)

-   load buildings dataset using `st_read` and the following SQL query to select only residential buildings\
-   hint: reproject data to EPSG:3083\

`SELECT *`  `FROM gis_osm_buildings_a_free_1`\
`WHERE (type IS NULL AND name IS NULL)`\
`OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')`\

```{r include=TRUE}
# define our query to select only residential buildings
query2 <- "SELECT*
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

# select residences from our buildings geopackage using our query and reproject to Texas Centric Albers Equal Area
houston_houses <- st_read('data/gis_osm_buildings_a_free_1.gpkg',
                          quiet = TRUE,
                          query = query2) %>%
  st_transform(crs = 3083)
```

##### find homes in blackout areas (20 points)

-   filter to homes within blackout areas\
-   count number of impacted homes\

```{r include=TRUE}
# filter to homes within blackout areas using spatial subsetting
blackout_houses <- houston_houses[houston_blackout, ]

# print the number of homes in blackout areas, building in a check to ensure that it is a subset
if (nrow(blackout_houses) < nrow(houston_houses)) {
  print(paste("Number of houston homes in blackout areas:", length(unique(blackout_houses$osm_id))))
} else {
  print("Check your subsetting method")
}
```

#### Investigate socioeconomic factors

##### load ACS data (10 points)

-   use `st_read()` to load the geodatabase layers\
-   geometries are stored in the `ACS_2019_5YR_TRACT_48_TEXAS` layer\
-   income data is stored in the `X19_INCOME` layer\
-   select the median income field `B19013e1`\
-   hint: reproject data to EPSG:3083\

```{r include=TRUE}
# read in Texas census tracts data and reproject to Texas Centric Albers Equal Area
texas_tracts <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", 
                  layer = "ACS_2019_5YR_TRACT_48_TEXAS", 
                  quiet = TRUE) %>% 
  st_transform(crs = 3083)

# read in the Texas census tracts tabular data and select the median income and GEOID columns
tracts_income <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", 
                  layer = "X19_INCOME", 
                  quiet = TRUE) %>% 
  dplyr::select(B19013e1, GEOID)
```

##### determine which census tracts experienced blackouts (10 points)

-   join the income data to the census tract geometries\
-   hint: make sure to join by geometry ID\
-   spatially join census tract data with buildings determined to be impacted by blackouts\
-   find which census tracts had blackouts

```{r}
# find the name of the column in texas_tracts that corresponds to GEOID in the tabular income data
# the function inside sapply returns a logical that tells whether any column in texas_tracts contains the value that matches the first GEOID value in tracts_income
names(texas_tracts)[sapply(texas_tracts, function(col) any(col %in% tracts_income$GEOID[1]))]
```

```{r include=TRUE}
# merge the census tracts and income data by common columns
texas_tracts <- merge(texas_tracts, tracts_income, by.x = "GEOID_Data", by.y = "GEOID")

# use st_intersects to find which census tracts experienced blackouts
# st_intersects returns a list of vectors, each vector containing the IDs of homes that intersect with that census tract
# using lengths returns the number of IDs in each vector, or the number of homes
# census tracts with a length > 0 experienced a blackout, assigning a "TRUE" to the blackout column
texas_tracts$blackout <- lengths(st_intersects(texas_tracts, blackout_houses)) > 0

# crop census tracts to the houston metro area
houston_tracts <- st_crop(texas_tracts, st_bbox(houston_blackout))

# print the number of impacted census tracts
print(paste("The number of Houston census tracts impacted by blackouts:",
            nrow(filter(houston_tracts, blackout == TRUE))))
```

##### compare incomes of impacted tracts to unimpacted tracts (10 points)

-   create a map of median income by census tract, designating which tracts had blackouts\
-   plot the distribution of income in impacted and unimpacted tracts
-   write approx. 100 words summarizing your results and discussing any limitations to this study

```{r}
# filter and subset for census tracts that experienced a blackout
blackout_tracts <- houston_tracts %>% 
  filter(blackout == TRUE)

# map census tracts by median income, highlighting census tracts that experienced a blackout
ggplot() +
  geom_sf(data = houston_tracts, # fill census tracts by median income and add a white border
          aes(fill = B19013e1),
          lwd = .1,
          col = "white") +
  scale_fill_viridis_b(name = "median income") + # add a name to the legend items
  geom_sf(data = blackout_tracts, # color borders of impacted tracts red
          aes(col = blackout),
          fill = "transparent",
          lwd = .2) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "transparent"),
                     breaks = "TRUE", # only include the symbol for impacted tracts in the legend
                     labels = c("experienced blackout"), # change the legend item label
                     name = "") +
  theme_void() + # remove latitude and longitude ticks
  labs(title = "Median Income and Blackouts by Census Tract in Houston") + # update the map title
  annotation_scale(width_hint = .4,
                   line_col = "white", 
                   text_col = "white",
                   pad_y = unit(.25, "in"),
                   pad_x = unit(.2, "in")) + # add a scale bar and change the color to white
  annotation_north_arrow(style = north_arrow_fancy_orienteering(line_col = "white", 
                                                                text_col = "white"),
                         pad_y = unit(.4, "in")) # add a north arrow and change the color to white
```

```{r}
# find the median median income by blackout status
medians <- houston_tracts %>% 
  group_by(blackout) %>% 
  summarize(median = median(B19013e1, na.rm = TRUE))

# visualize the distribution of income in impacted versus unimpacted census tracts using a boxplot
ggplot(houston_tracts, aes(x = B19013e1, y = blackout, fill = blackout)) +
  geom_boxplot() +
  geom_text(data = medians, # add median labels for each group
            aes(x = median, y = blackout, label = median), 
            size = 3, 
            vjust = -.5, # move the label off of the line
            angle = -90) + # rotate the label
  theme_bw() +
  theme(legend.position = "none") + # remove the legend
  labs(title = "Distribution of Median Income in Census Tracts by Blackout Status",
       x = "median income",
       y = "blackout status") + # update the title and x and y axes labels
  scale_y_discrete(labels = c("no blackout", "blackout")) # update the blackout group labels
```

-   The box plot displaying the distribution of median income in impacted and unimpacted census tracts shows that impacted census tracts actually had a slightly higher median median income than unimpacted census tracts. The difference is small, with impacted census tracts showing a median median income of \$ `r medians$median[2]` versus unimpacted census tracts showing a median median income of \$ `r medians$median[1]` . While we might expect to see a lower median median income in impacted census tracts, that isn't the case. A big limitation of our study could be the availability and accuracy of our light intensity rasters. The rasters appeared to have noise that could have affected the creation of our blackout mask, especially since we manually assigned the blackout threshold to be a difference of 200 or more nW cm^-2^sr^-1^ . Without accounting for noise, we may have misrepresented blackout areas.
